---
title: "Permisos Proyecto"
author: "Rebecca Stamm"
date: "2025-06-18"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true

---

## Set-Up

```{r}
suppressMessages(library(tidyverse))
library(ggplot2)
library(readxl)

# reading in files
permisos <- read_excel("permisos.xls")
permisos19 <- read_excel("PC Monteverde 2019.xls")
permisos22 <- read_excel("PC Monteverde 2022.xls")
# using the same names for each columns
names(permisos) = names(permisos22)
# merging the separate data frames together
all_permisos <- rbind(permisos19, permisos, permisos22)

# to type out accents in graphs or have the months in the dates in spanish
# Sys.setlocale("LC_MESSAGES", "es_ES.UTF-8")
```


## EDA

```{r eval=FALSE, include=FALSE}
summary(all_permisos$fecha_pago)
unique(all_permisos$tipo_obra)
# table(all_permisos$propietarioa[duplicated(all_permisos$propietarioa)])
table(all_permisos$propietarioa[(all_permisos$propietarioa != all_permisos$solicitante)])
```

### Permit Type EDA
```{r}
all_permisos$destino <- factor(all_permisos$destino)

# bar graph
all_permisos |>
  ggplot(aes(x = fct_infreq(destino))) +
  geom_bar(fill = "darkgreen") +
  theme(axis.text.x = element_text(angle = 50, hjust = 1)) +
  geom_text(stat = "count", aes(label = after_stat(count)),
            vjust = -0.5) +
  labs(title = "Los Destinos de los Permisos / Permit Uses", 
       x = "Destino / Use", y = "Conteo / Count") +
  scale_y_continuous(limits = c(0, 310))

# table version
all_permisos %>% group_by(destino) %>% summarize(n())
```

### Building Material EDA
```{r}
# Walls
all_permisos |>
  ggplot(aes(x = fct_infreq(paredes_exteriores))) +
  geom_bar() +
  theme(axis.text.x = element_text(angle = 50, hjust = 1)) +
  geom_text(stat = "count", aes(label = after_stat(count)),
            vjust = -0.5) +
  labs(title = "Las Paredes Exteriores de los Permisos", 
       x = "Material de las Paredes Exteriores",
       y = "Conteo") +
  scale_y_continuous(limits = c(0, 85))

# Floors
all_permisos |>
  ggplot(aes(x = fct_infreq(piso))) +
  geom_bar() +
  theme(axis.text.x = element_text(angle = 50, hjust = 1)) +
  geom_text(stat = "count", aes(label = after_stat(count)),
            vjust = -0.5) +
  labs(title = "Los Pisos de los Permisos", 
       x = "Material de los Pisos",
       y = "Conteo") +
  scale_y_continuous(limits = c(0, 150))

# Roofs
all_permisos |>
  ggplot(aes(x = fct_infreq(techo))) +
  geom_bar() +
  theme(axis.text.x = element_text(angle = 50, hjust = 1)) +
  geom_text(stat = "count", aes(label = after_stat(count)),
            vjust = -0.5) +
  labs(title = "Los Techos de los Permisos", 
       x = "Material de los Techos",
       y = "Conteo") +
  scale_y_continuous(limits = c(0, 300))

# aes(fill = destino)  # option to add a third variable to these graphs
```

### Text Processing
What words appear the most in `direccion`?
```{r}
p <- paste(all_permisos$direccion)
s <- unlist(strsplit(p,split=" |[[:punct:]]"))
sort(table(s),decreasing=TRUE)[1:30]
```

```{r}
library(tidytext)
all_permisos %>% select(.,direccion) %>%
  unnest_tokens(.,bigram,direccion,token="ngrams",n=2) -> all_permisos.new
sort(table(all_permisos.new$bigram),decreasing=TRUE)[1:50]
```


## Municipalities

### Variable Creation
```{r}
# municipality variable creation
# word processing with address

all_permisos %>% filter(.,str_detect(direccion,"LOS LLANOS")) %>% nrow(.)
all_permisos %>% filter(.,str_detect(direccion,"SAN LUIS")) %>% nrow(.)
all_permisos %>% filter(.,str_detect(direccion,"MONTEVERDE")&
                          !str_detect(direccion,"SAN LUIS")) %>% nrow(.)
all_permisos %>% filter(.,str_detect(direccion,"SANTA ELENA")) %>% nrow(.)
all_permisos %>% filter(.,str_detect(direccion,"BARRIO LA VIRGEN")) %>% nrow(.)
all_permisos %>% filter(.,str_detect(direccion,"CERRO PLANO")) %>% nrow(.)
all_permisos %>% filter(.,str_detect(direccion,"VALLE BONITO")) %>% nrow(.)
all_permisos %>% filter(.,str_detect(direccion,"LA LINDORA")) %>% nrow(.)
all_permisos %>% filter(.,str_detect(direccion,"SAN LUIS")) %>% nrow(.)

muni <- rep(NA,nrow(all_permisos))
muni[all_permisos %>% with(.,which(str_detect(direccion,"LOS LLANOS")))] <- 
  "LOS LLANOS"
muni[all_permisos %>% with(.,which(str_detect(direccion,"SAN LUIS")))] <- 
  "SAN LUIS"
muni[all_permisos %>% with(.,which(str_detect(direccion,"MONTEVERDE")&
                                     !str_detect(direccion,"SAN LUIS")))] <- 
  "MONTEVERDE"
muni[all_permisos %>% with(.,which(str_detect(direccion,"SANTA ELENA")))] <- 
  "SANTA ELENA"
muni[all_permisos %>% with(.,which(str_detect(direccion,"BARRIO LA VIRGEN")))] <- 
  "BARRIO LA VIRGEN"
muni[all_permisos %>% with(.,which(str_detect(direccion,"CERRO PLANO")))] <- 
  "CERRO PLANO"
muni[all_permisos %>% with(.,which(str_detect(direccion,"VALLE BONITO")))] <- 
  "VALLE BONITO"
muni[all_permisos %>% with(.,which(str_detect(direccion,"LA LINDORA")))] <- 
  "LA LINDORA"
all_permisos <- cbind(all_permisos, muni)
```

### Municipality EDA
```{r}
# Municipality vs. Valuation
all_permisos %>% ggplot(.,aes(x=muni,y=log10(valor_de_la_obra))) + 
  geom_boxplot(fill="mediumorchid") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + 
  xlab(NULL) + ylab("Log(Valuation)")

# Floor Material by Municipality
ggplot(all_permisos,aes(fill=muni,x=fct_infreq(piso))) + 
  geom_bar(position="fill") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

```{r}
# Number of Floors by Municipality (proportional)
ggplot(all_permisos,aes(fill=muni,x=numero_pisos)) + geom_bar(position="fill") 

# Number of Bedrooms by Municipality (stacked)
ggplot(all_permisos,aes(fill=muni,x=numero_dormitorios)) +
  geom_bar(position="stack")
```

### Destino vs. Municipality Bar Charts
```{r}
# Permit Types by Municipality (proportional)
ggplot(all_permisos, aes(fill=muni,x=fct_infreq(destino))) + 
  geom_bar(position="fill") +
  theme(axis.text.x = element_text(angle = 50, hjust=1)) +
  labs(x = "Destino / Use", y = "Proporcion / Proportion", 
       fill = "Municipio / Municipality",
       title = "Los Destinos de los Permisos por Municipio /
                Permit Uses by Municipality")

# Municipality by Permit Types (stacked)
ggplot(all_permisos, aes(fill=destino,x=fct_infreq(muni))) + 
  geom_bar(position="stack") +
  theme(axis.text.x = element_text(angle = 50, hjust=1)) +
  labs(x = "Municipio", y = "Conteo", fill = "Destino")
```


## Spatial Distribution of Permits 

### Coordinate Conversion
```{r}
library(sf)  # necessary for the functions used to transform the coordinates

# relabeling an NA row of coordinates to (0,0) to represent NA
all_permisos$Coordenadas.X[all_permisos$n_permiso == 1749] <- 0
all_permisos$Coordenadas.Y[all_permisos$n_permiso == 1749] <- 0

# converting UTM (Universal Transverse Mercator) to latitude/longitude
# Coordenadas.X is the easting value
# Coordenadas.Y is the northing value

convert_coords <- function(df, x_col = "x", y_col = "y") {
  # Projection Name: UTM Zone 16N (WGS84)
  # note: Costa Rica spans UTM Zones 16N and 17N, but Monteverde is in Zone 16N
  point_utm <- st_as_sf(df, coords = c(x_col, y_col), crs = 32616)
  point_wgs84 <- st_transform(point_utm, crs = 4326)
  coords <- st_coordinates(point_wgs84)
  
  # rewriting to dataframe
  df$new_lon <- coords[, 1] + 3 # the +3 was just the only thing i got to work
  df$new_lat <- coords[, 2] 
  
  return(df)
}

# applying the function to the dataset
all_permisos <- convert_coords(all_permisos, "Coordenadas.X", "Coordenadas.Y")
# checking that it worked
head(all_permisos$new_lon) # should be in the -84.8 range
head(all_permisos$new_lat) # should be in the  10.3 range

# filter out the zeros (could also do this before, but i didn't remember to)
all_permisos_filter <- all_permisos %>% 
  filter(!new_lon == -88.48874388 & !new_lat == 0)
```

### Factoring Variables for the Map
```{r}
# factoring destino into fewer groups for the map legend to not be cluttered
all_permisos_filter <- all_permisos_filter %>%
  mutate(destino_grouped = fct_lump_min(destino, min = 3, # groups w/ < 3
                                        other_level = "OTROS")) # other

all_permisos_filter %>% group_by(destino_grouped) %>% summarize(n())

# what data was lost in removing the NA coordinates
all_permisos_filter |>
  ggplot(aes(x = fct_infreq(destino_grouped))) +
  geom_bar(fill = "darkgreen") +
  theme(axis.text.x = element_text(angle = 50, hjust = 1)) +
  geom_text(stat = "count", aes(label = after_stat(count)),
            vjust = -0.5) +
  labs(title = "Los Destinos de los Permisos", x = "Destino", y = "Conteo") +
  scale_y_continuous(limits = c(0, 310))

# adding a year variable to separate the map as well
all_permisos_filter$year <- year(all_permisos_filter$fecha_pago)
all_permisos_filter$year <- factor(all_permisos_filter$year)
```

### Mapping
```{r}
# # with google api map
# library(ggmap)
# register_google(key = "AIzaSyDjXPJxSS4LNqUxq1EnxzPYMuEBUYhNzic", write = TRUE)

# center_lon <- mean(range(all_permisos_filter$new_lon, na.rm = TRUE))
# center_lat <- mean(range(all_permisos_filter$new_lat, na.rm = TRUE))

# monteverde_map <- get_map(location = c(lon = center_lon, lat = center_lat),
#                           zoom = 10,
#                           source = "google",
#                           maptype = "terrain")
# 
# ggmap(monteverde_map) +
#   geom_point(data = all_permisos_filter, aes(x = new_lon, y = new_lat), 
#              alpha = 0.5)

# the above code works, but the range of points is too large to fit on the 
# static google api map, so i switched to the leaflet package below

# interactive leaflet map
library(leaflet)

# setting the destino point coloring
pal <- colorFactor(palette = c("#000000", "#009E73", "#0072B2",
                               "#E69F00", "#CC79A7", "#F0E442"),
                   domain = all_permisos_filter$destino_grouped)

leaflet(data = all_permisos_filter) %>%
  addTiles() %>%
  addCircleMarkers(~new_lon, ~new_lat, 
                   radius = 7,              # points enlarged for presentation
                   color = ~pal(destino_grouped), # colored by destino
                   stroke = FALSE,
                   fillOpacity = 0.5,
                   group = ~year,           # sectioned by year (toggle)
                   # when click on point, address appears
                   popup = ~paste("Direccion: ", direccion,
                                  "<br>Fecha: ", fecha_pago)) %>% 
  addLegend("bottomright",
            pal = pal,
            values = ~destino_grouped,
            title = "Destino") %>% 
  addLayersControl( # adds the overlay that allows to toggle between years
    overlayGroups = unique(all_permisos_filter$year),
    options = layersControlOptions(collapsed = FALSE)
  )
```

```{r}
# another map with points colored by year

pal2 <- colorFactor(palette = c("#009E73", "#0072B2","#CC79A7","#F0E442"),
                    domain = all_permisos_filter$year)

leaflet(data = all_permisos_filter) %>%
  addTiles() %>%
  addCircleMarkers(~new_lon, ~new_lat, 
                   radius = 6, 
                   color = ~pal2(year),
                   stroke = FALSE,
                   fillOpacity = 0.5,
                   popup = ~paste("Direccion: ", direccion,
                                  "<br>Fecha: ", fecha_pago)) %>% 
  addLegend("bottomright",
            pal = pal2,
            values = ~year,
            title = "Ano")
```


## The Rate of Construction

### Number of Permits by Month
```{r}
monthly_permits <- all_permisos %>%
  mutate(month = floor_date(fecha_pago, "month")) %>%
  group_by(month) %>%
  summarise(total = n())

ggplot(monthly_permits, aes(x = as.Date(month), y = total)) +
  geom_line() +
  geom_smooth(se = TRUE, span = 0.7) +
  # scale_x_date(date_breaks = "3 months",
  #              date_labels = "%b %Y") +  # adds months to x-axis
  labs(title = "Permisos por Mes / Permits by Month",
       x = "Fecha / Date", y = "Numero de Permisos / Number of Permits") +
  # theme(axis.text.x = element_text(angle = 50, hjust=1)) +  # if x in months
  geom_vline(color = "red", linewidth = 1, linetype = "dashed", 
             xintercept = as.Date("2020-3-6")) +
  geom_text(label = "Pandemia de\nCOVID-19", x = as.Date("2020-6-15"), y = 19)
```

### Autocorrelations
```{r}
# By day
daily_permits <- all_permisos %>% 
  count(fecha_pago) %>% 
  arrange(fecha_pago)

permit_ts <- ts(daily_permits$n, frequency = 365)
acf(permit_ts, main = "Autocorrelation of Daily Permits")

# By month
monthly_permits <- all_permisos %>% 
  mutate(month = floor_date(fecha_pago, "month")) %>% 
  count(month) %>% 
  arrange(month)

permit_ts_mes <- ts(monthly_permits$n, start = c(2019, 08, 01), frequency = 12)
acf(permit_ts_mes, main = "Autocorrelation of Monthly Permits")
```

### Time Series Decomposition
```{r}
# decomposition
permit_ts_mes <- ts(monthly_permits$n, start = c(2019, 08, 01), frequency = 12)
decomp <- stl(permit_ts_mes, s.window = "periodic")
plot(decomp)

# seasonality strength
## how much variation in the data is due to the seasonality (closer to 1, better)
var(decomp$time.series[, "seasonal"]) /
  var(decomp$time.series[, "seasonal"] + decomp$time.series[, "remainder"])

# trend strength
## how much variation in the data is due to the overall trend
var(decomp$time.series[, "trend"]) /
  var(decomp$time.series[, "trend"] + decomp$time.series[, "remainder"])

# seeing the seasonal effect by month to better understand the tsd
seasonal <- decomp$time.series[, "seasonal"]
months <- cycle(permit_ts_mes)
seasonal_avg <- tapply(seasonal, months, mean)
seasonal_df <- data.frame(
  Month = month.name,
  SeasonalEffect = as.numeric(seasonal_avg)
)

ggplot(seasonal_df, aes(x = reorder(Month, SeasonalEffect), y = SeasonalEffect)) +
  geom_col(fill = "steelblue") +
  coord_flip()
```

### All Permisos by Month (Regardless of Year)
```{r}
all_permisos %>%
  mutate(month = month(fecha_pago, label = TRUE)) %>%
  count(month) %>%
  ggplot(aes(x = month, y = n)) +
  geom_col(fill = "steelblue") +
  labs(x = "mes", y = "conteo", title = "Permisos por Mes")
```

### Heatmap of Monthly Permit Count
```{r}
heatmap_data <- all_permisos %>%
  mutate(year = year(fecha_pago),
         month = month(fecha_pago, label = TRUE)) %>%
  group_by(year, month) %>%
  summarise(count = n(), .groups = "drop")

ggplot(heatmap_data, aes(x = month, y = factor(year), fill = count)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "darkblue", high = "darkorange") +
  labs(title = "Heatmap de Permisos / Permit Heatmap",
       x = "Mes / Month", y = "Año / Year") 
```

### Monthly Permit Count by Municipality 
Does it differ by municipality? no
```{r}
permisos_por_muni <- all_permisos %>%
  mutate(month = floor_date(fecha_pago, "month")) %>%
  group_by(month, muni) %>% 
  summarise(total = n(), .groups = "drop")

permisos_por_muni |>
  ggplot(aes(x = month, y = total)) +
  geom_line() +
  facet_wrap(~muni) +
  labs(title = "Permisos Mensuales por Municipio",
       x = "Mes", y = "Conteo")

heatmap_muni <- all_permisos %>%
  mutate(year = year(fecha_pago),
         month = month(fecha_pago, label = TRUE)) %>%
  group_by(muni, year, month) %>%
  summarise(total = n(), .groups = "drop")

heatmap_muni |>
  ggplot(aes(x = month, y = factor(year), fill = total)) +
  geom_tile(color = "white") +
  facet_wrap(~muni) +
  scale_fill_gradient(low = "darkblue", high = "darkorange") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + 
  labs(title = "Heatmap de Permisos por Municipio",
       x = "Mes", y = "Year") 
```

### Average Monthly Permit Count by Municipality
```{r}
all_permisos %>%
  group_by(muni) %>%
  summarise(mean_per_month = n() / 
              n_distinct(floor_date(fecha_pago, "month")))
```

### Monthly Permit Count by Type
```{r}
permisos_por_destino <- all_permisos %>% 
  mutate(month = floor_date(fecha_pago, "month")) %>% 
  group_by(month, destino) %>% 
  summarise(total = n(), .groups = "drop")

permisos_por_destino |>
  ggplot(aes(x = month, y = total, fill = destino)) +
  geom_col() +
  labs(title = "Permisos Mensuales por Destino",
       x = "Mes", y = "Conteo", fill = "Destino")

permisos_por_destino |>
  ggplot(aes(x = month, y = total, color = destino)) +
  geom_line()
```
Number of vacant lots has increased recently, maybe something to note.

### Hypothesis Testing
```{r}
agg <- all_permisos %>% 
  mutate(month = floor_date(fecha_pago, "month")) %>% 
  count(muni, month)

summary(aov(n~muni, data = agg))

table <- table(all_permisos$muni, all_permisos$destino)
chisq.test(table)
```
